---
title: "Global Plastic Waste Report"
author: "Dhruvin Modh, Sandhya Erland Chandrasekar, Xinyu Wang"
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    theme: metropolis
    echo: false
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    theme: metropolis
    echo: false
editor: visual
---

## Executive Summary

This report investigates global plastic waste patterns, focusing on 165 countries with varying recycling practices and environmental risks. We explore how per capita plastic waste and national recycling rates relate to coastal pollution vulnerability. Using figures and tables, we highlight the countries with the most plastic intensity and evaluate systemic differences. The findings offer useful guidance for policy improvements and environmental planning worldwide.

## Introduction

Plastic waste is a growing global concern, particularly for countries near coastlines. Poor waste management not only pollutes land ecosystems but also threatens marine environments. This report aims to analyze patterns in plastic waste production, recycling performance, and coastal pollution risk using a recent global dataset. The dataset includes 165 countries and features metrics such as total plastic waste, per capita generation, and recycling rates.

By comparing countries based on their waste intensity and risk levels, we seek to understand which factors contribute to environmental vulnerability. We use both descriptive and comparative methods, and summarize our findings through visualizations and aggregated statistics. Our focus is on highlighting regional disparities and identifying outliers.

The purpose of this analysis is to provide actionable insights for decision-makers in environmental and public policy sectors. The report contributes to a broader understanding of how waste systems interact with geography, policy, and infrastructure.


## **Methodology**

## 

```{r}
library(tidyverse)
library(janitor)
library(knitr)

```

The analysis was conducted using the “Plastic Waste Around the World” dataset, which contains country-level data on plastic waste generation, recycling rates, and coastal vulnerability. The dataset was imported from a GitHub source and processed in R using reproducible and version-controlled workflows.

Upon loading the dataset, we conducted several preprocessing steps:

1\. **Variable Cleaning and Transformation:** All column names were standardized using \`janitor::clean_names()\`. Variables with missing values, particularly \`per_capita_waste_kg\` and \`coastal_waste_risk\`, were excluded from analysis. The continuous variable \`recycling_rate\` was binned into three categories: "Low", "Medium", and "High" recycling performance, using quantile-based thresholds.

2\. **Derived Metrics**: We introduced a new variable, \`total_waste_per_person_mt\`, calculated by dividing \`total_plastic_waste_mt\` by an estimated population derived from total waste and per capita rates. This derived variable enables us to assess plastic production intensity across countries more consistently.

3\. **Risk-based Grouping**: Countries were grouped by \`coastal_waste_risk\` (Low, Medium, High) and stratified by binned recycling performance. This allowed for a multifaceted analysis of waste management practices in different geographic and policy contexts.

```{r}

# Load and clean
plastic_data <- read_csv("Data/Plastic Waste Around the World.csv") %>%
  clean_names() %>%
  filter(!is.na(per_capita_waste_kg), !is.na(coastal_waste_risk), !is.na(recycling_rate))

# Bin recycling rates
plastic_data <- plastic_data %>%
  mutate(
    recycling_category = ntile(recycling_rate, 3),
    recycling_category = case_when(
      recycling_category == 1 ~ "Low",
      recycling_category == 2 ~ "Medium",
      recycling_category == 3 ~ "High"
    )
  )

# Derived variable: total waste per person (in MT)
plastic_data <- plastic_data %>%
  mutate(
    est_population_million = total_plastic_waste_mt * 1e6 / per_capita_waste_kg,
    waste_per_person_mt = total_plastic_waste_mt / est_population_million
  )
```

4\. **Visualization and Tabulation**: We created two key outputs:

Figure 1 : A bar chart of the top 10 countries by per capita plastic waste, with reordered factors and enhanced labels.

Table 1 : A grouped summary table showing the mean and standard deviation of plastic waste variables stratified by coastal risk level.

```{r}
top10_enhanced <- plastic_data %>%
  arrange(desc(per_capita_waste_kg)) %>%
  slice(1:10)

ggplot(top10_enhanced, aes(x = reorder(country, per_capita_waste_kg), y = per_capita_waste_kg, fill = recycling_category)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Countries by Per Capita Plastic Waste",
    x = "Country",
    y = "Plastic Waste (kg per person)",
    fill = "Recycling Category"
  ) +
  theme_minimal()

```

```{r}
summary_table <- plastic_data %>%
  group_by(coastal_waste_risk) %>%
  summarise(
    Mean_Total_Waste_MT = mean(total_plastic_waste_mt, na.rm = TRUE),
    SD_Total_Waste_MT = sd(total_plastic_waste_mt, na.rm = TRUE),
    Mean_Per_Capita_KG = mean(per_capita_waste_kg, na.rm = TRUE),
    SD_Per_Capita_KG = sd(per_capita_waste_kg, na.rm = TRUE),
    Mean_Recycling = mean(recycling_rate, na.rm = TRUE)
  )

kable(summary_table, digits = 2)

```


# Results
```{r}
#| label:  tbl-summarytable
#| tbl-cap: "Plastic Waste Indicators by Coastal Risk Level"
#| echo: true
#| warning: false
#| message: false

library(readr)
library(dplyr)
library(janitor)
library(knitr)

# Read the CSV file using a relative path
plastic_data <- read_csv("data/Plastic Waste Around the World.csv")

# Clean column names (convert to lower case with underscores)
plastic_data <- clean_names(plastic_data)

# Group by coastal risk level and calculate summary statistics
summarytable <- plastic_data %>%
  group_by(coastal_waste_risk) %>%
  summarise(
    `Per_Capita` = mean(per_capita_waste_kg, na.rm = TRUE),
    `Total_Capita` = mean(total_plastic_waste_mt, na.rm = TRUE),
    `SD(Total)` = sd(total_plastic_waste_mt, na.rm = TRUE),
    `Mean Recycling(%)` = mean(recycling_rate, na.rm = TRUE)
  ) %>%
  arrange(match(coastal_waste_risk, c("Very_High", "High", "Medium", "Low")))  

# Display the table with two decimal places
kable(summarytable, digits = 2,  align = "c")


```
As shown in @tbl-summarytable, plastic waste data from 165 countries is grouped by coastal risk level.

Very High risk countries have the lowest waste per person (33.68 kg), but they still produce the second-largest total waste (2.52 million tonnes), likely due to large populations.Medium risk countries have the highest plastic waste per person (163.46 kg) and also the highest total waste (2.94 million tonnes). This means people in these countries use a lot of plastic, and the overall burden is also high.High risk countries show medium levels of waste per person (89.67 kg), but they have the lowest average recycling rate (5.50%) and the largest difference in total waste between countries (standard deviation of 7.44).Low risk countries produce the smallest amount of total waste (0.97 million tonnes), have more similar waste levels across countries (standard deviation of 1.30), and the highest recycling rate (19.66%).

This shows that coastal waste risk is not simply based on how much waste each person produces. Other factors, like population size and waste system quality, also play an important role.  


